Nonstationary, Quasi-Periodic Risk of Climate Extremes, and Decision Analysis

### 2018-02-21 writeup/main.tex
add slurm

\input{tex/header/macros.tex}
\input{tex/header/packages.tex}

\title{Climate Change Adaptation: Uncertainty Associated with Sequential Finite Period Decisions under Nonstationarity}
\author[1,2]{James Doss-Gollin}
\author[1,2]{Upmanu Lall}
\author[1,2]{David Farnham}
\affil[1]{Columbia Water Center, Columbia University}
\affil[2]{Department of Earth and Environmental Engineering, Columbia University}

\input{tex/header/biblatex.tex}
\addbibresource{../library.bib}

@@

\begin{document}
\maketitle
\RaggedRight{}
\begin{abstract}
  Hydroclimatic systems exhibit organized low-frequency and regime-like variability at multiple time scales, causing the risk associated with climate extremes such as floods and droughts to vary at multiple timescales.
  Despite broad recognition of this nonstationarity, there has been little theoretical development of ideas for the design and operation of infrastructure that simultaneously consider the potential for predictability of secular change and regime behavior or low-frequency variability.
  Consequently, existing theories of flood risk assessment are well-suited to analysis of flood adaptation strategies with fixed project operation period \(M\), but not for analysis of sequential decision strategies.
  In this paper we illustrate some basic considerations of uncertainty and risk analysis associated with the sequential decisions that may need to be made for climate risk adaptation in a nonstationary adaptation.
  We simulate synthetic data and fit the resulting simulations in a Bayesian context using stationary and non-stationary models to show that \(\ldots{}\).
  This is a first conceptual step to the development of a full sequential decision analysis approach under nonstationarity.
\end{abstract}

@@

Hydroclimatic systems vary on many timescales, including regime-like variability and secular variability  \citep{Milly2008,Merz2014,Hurst1951,Sveinsson2005,Hodgkins2017,Matalas2012}.
This multi-scale variability greatly complicates the estimation of future flood sequences, leading to ``low confidence in projections of changes in fluvial floods'' \citep{IPCC2012}.
Yet the same memory and low-frequency variability that complicate projecting flood sequences far into the future may also impart some predictability at short timescales \citep{Jain2001}.
The existence of this information gap suggests that approaches for adapting to hydroclimate risks, such as floods, which have short project operation periods, may be preferential in some cases to large and permanent projects with long project operations.

At present, most approaches to nonstationary flood risk management, or climate change adaptation more broadly, have framed the problem first as one of detecting historical change or modeling future change, and then assessing the implications of this change for future infrastructure.
Though more recent studies have pointed out that the paradigm of requiring proof of the existence of trends before adapting to nonstationarity does not lead to optimal decisionmaking \citep{Vogel2013,Rosner2014}, such discussion has also focused on linear time trends and on choosing whether or not to implement a particular adaptation policy with a fixed planning period.
In contrast, public and private actors seeking to adapt to changing flood or climate risk typically have access to a suite of options including large flood control infrastructure, insurance contracts and other financial instruments, vulnerability reduction (\ie{} disaster response training), and exposure reduction (\ie{} moving infrastructure out of the floodplain).
In this context, understanding under what conditions a large, permanent project or small, easily modified adaptation measure are optimal becomes an important question.

In this paper we provide some initial insight into the bias and uncertainty associated with flood frequency estimation for projects as a function of the project planning period \(M\) and the length of the available data record \(N\).
This project planning period can be interpreted as the design life of a physical structure, or as the contract length of a financial instrument.
Our aim is not to identify the ``best'' model, but rather to build some theory as to how the underlying nature of the natural variability and nonstationarity and the choice of model used to model future flood risk map onto estimation bias and variance in the context of many different \(M\) and \(N\).
This estimation bias and variance may map directly onto the price of an insurance contract\footnote{cite} or the possibility of over/under-design of a flood control structure.
Though we focus on flood risk, this type of analysis may be used for understanding the implications of uncertain projections of the future for climate hazards including drought, hurricane, wind, \etc{}.

@@

We use an idealized model to generate sequences of annual-maximum floods, fit these sequences in a probabilistic framework to three flood models that incorporate time in different ways, and evalaluate the performance of these models for multiple observational record lengths \(N\) and project planning periods \(M\).

We generate synthetic flood sequences from two different simple conceptual models that represent chaotic dynamics and long memory.

For both cases, the expected value of the annual maximum flood flood follows a log-normal distribution:
\begin{equation} \label{eq:lognormal}
  \log Q(t) \big| \mu(t),\sigma(t) \sim \normal \qty(\mu(t), \sigma(t)).
\end{equation}
where the parameters \( \mu(t), \sigma(t) \) are allowed to vary in time following the two approaches outlined below.
For simplicitiy we assume a constant coefficient of variation:
\begin{equation}
  \sigma(t) = \begin{cases} \alpha \mu(t) & \qqtext{if} \alpha \mu(t) \geq \sigma_\text{min} \\ \sigma_\text{min} & \qqtext{else} \end{cases}
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.
The two models are summarized in \cref{tab:model-generating}.

Though both approaches used to generate sequences \( \mu(t) \) useh highly simplified models neglecting many important mechanisms, they enable us to generate many synthetic sequences of \(\qty[M + N]\) years which exhibit chaotic but organized quasi-periodic behavior (regime behavior or low-frequency variability) plus a trend term, thereby capturing some essential features of hydrological variability.

In \cref{sec:sequence-realistic} we show sequences of streamflow generated by both models for some choices of parameters and discuss the processes which they can reasonably represent.

\begin{table}[bht]
  \begin{center}
    \begin{tabular}{L{1.0in} L{3.0in} L{1.5in}}
      \toprule
        Model Name & Description & Relevant Section \\
      \midrule
        NINO3 & \(\mu(t)\) proportional to synthetic NINO3 time series & \cref{sec:methods-enso} \\
        Markov & \(\mu(t)\) depends only on sequence of latent states generated by Markov chain & \cref{sec:methods-markov} \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{Summary of models used for generating synthetic flood sequences\label{tab:model-generating}}
\end{table}

\subsubsection{Synthetic ENSO Model\label{sec:methods-enso}}

ENSO is the dominant mode of tropical climate variability, impacts flood risk around the world, and varies in a quasi-oscillatory manner with variability on the order of \SIrange{4}{6}{year} modulated by low-frequency oscillation \citep[see][for a comprehensive review]{Sarachik2010}.
A \SI{100000}{year} integration of the Cane-Zebiak model \citep{Zebiak1987} was used to produce a monthly NINO3.4 index with stationary forcing as described in \citet{Ramesh2017}.
A \SI{1000}{year} subset of this data is shown in \cref{fig:enso-ts}.
To create an annual time series, we average the October-December months to create a time series \( x(t) \) where \( t \) is taken in annual steps \(1, \ldots, \num{100000} \).

First a sequence of length \(M+N\) is randomly chosen from the \SI{100000}{year} ENSO sequence.
Next, \(\mu\) is calculated following by
\begin{equation}
  \mu(t) = \mu_0 + \beta x(t) + \gamma \qty(t - t_0)
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.

We do not, in general, know the true distribution of annual-maximum floods, and as discussed in \cref{sec:ffa} a plethora of approaches have been suggested in the literature.
This paper does not propose new models or suggest a general ``best'' model; instead it aims to describe the total uncertainty in estimates using easily methods which have been well studied in the literature.

In general, each historical \(N\)-year sequence of streamflow is fit to a statistical model, from which 1000\footnote{Check this parameter} sequences of future streamflow are generated using Monte Carlo simulation.
These estimates are then used to calculate \( \hat{p_T} \), and the estimates are evaluated as described in \cref{sec:methods-uncertainty}.
Three models are assessed for estimating future streamflow; they are summarized in \cref{tab:model-fitting} and described in greater detail below.

\begin{table}[bht]
  \begin{center}
    \begin{tabular}{L{1.25in} L{3.0in} L{1.25in}}
      \toprule
        Model Name & Description & Relevant Section \\
      \midrule
        \texttt{ln2\_stationary} & \( \log Q \sim N \qty(\mu, \sigma) \) & \cref{sec:method-stationary} \\
        \texttt{ln2\_trend} & \(\log Q \sim N \qty(\mu_0 + \beta_\mu \qty[t - t_0], \sigma_0 + \beta_\sigma \qty[t - t_0] ) \) & \cref{sec:method-trend} \\
        \texttt{HMM} & Hidden Markov Model & \cref{sec:method-HMM} \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{Summary of models used for fitting\label{tab:model-fitting}}
\end{table}

\subsubsection{Stationary Model\label{sec:method-stationary}}

We do not, in general, know the true distribution of annual-maximum floods, and there has been substantial discussion in the literature as to the choice of distribution to use for stationary flood frequency analysis.
We consider the log-normal model (\cref{eq:lognormal}) for stationary flood distribution; while this model is less flexible than three-parameter models such as the GEV or LP3 \citep{Vogel1996}, it is computationally convenient and easy to understand.
More importantly, since we know the true distribution of these synthetically generated flood, using the log-normal for both generating and fitting supports interpreting results as a lower bound on total uncertainty.
In the stationary context, we assume \(\mu(t) = \mu\) and \(\sigma(t)=\sigma\).

For each simulation, the first \(N\) years \(\qty(t=[t_0-N+1, \ldots, t_0])\) are treated as observations.
Then, once the appropriate time parameterization is selected (\cref{tab:model-fitting}), the parameters are fit in a Bayesian framework to fully represent the posterior uncertainty as implemented in the \texttt{stan} probabilistic programming language \citep{Carpenter2016}.
Weak (uninformative) priors are chosen; in the context of a synthetic data experiment using informative priors seems inappropriate but other studies\footnote{cite} have used scaling information and regional information to inform prior distributions and reduce posterior variance.

\subsubsection{Trend Model\label{sec:method-trend}}

Many studies\footnote{cite} have used time parameterizations for fitting.
We consider a trend model where the data is log-normally distributed as \cref{eq:lognormal} with a linear time trend on both \( \mu \) and \( \sigma \):
\begin{align}
  \mu(t) = \mu_0 + \beta_\mu \qty(t - t_0) \\
  \sigma(t) = \sigma_0 + \beta_\sigma \qty(t - t_0)
\end{align}

As for the stationary case, we fit posterior samples in a Bayesian framework using \texttt{stan} and apply weak priors.
The influence of priors in this case is stronger than in the stationary case, and a reasonable choice of priors might strongly favor \(\beta_mu=0\) and \(\beta_\sigma=0\), for example using the Horseshoe prior parameterization \citep{Piironen2016a}.

A Hidden Markov Model (HMM) is a statistical Markov model in which the system being modeled is assumed to be a Markov process with unobserved (\ie{} hidden) states \citep{Rabiner1986}.
The (unobserved) states evolve following a first-order Markov process, and the observed variable (\ie{} streamflow) depends only on the observed state.
As such, HMMs are a structured prediction method which extend general mixture models to sequences of data, where position in the sequence is relevant.
Because this structure allows the hidden state to represent hydroclimatological ``regimes'' \citep{Reinhold1982,Michelangeli1995,Merz2014}, HMMs have have been used extensively to model hydrological variables such as streamflow \citep{Bracken2016}.
HMMs have also been used to model the evolution of climate states such as ENSO.\footnote{Julian's work}

To supplement the distributional models described in \cref{sec:estimation}, we also fit observed streamflow sequences using a HMM as implemented in the python \texttt{pomegranate} package \citep{Schreiber2016}.
The model is fit to the \(N\) observed years of data using the Baum-Welch algorithm, assuming that the data follow a log-normal distribution conditional only on the state.
To choose the number of hidden states, a log probability score is used.
Future floods are estimated by simulating future states from the resulting matrix and then drawing floods log-normally using the estimated parameters.

\texttt{How much more do we need to say about HMM?}

\subsection{Reproducability}

\begin{figure}[b]
  \includegraphics[width=\textwidth]{../../figs/stationary_sequences.pdf}  
  \caption{
    50 streamflow sequences from (Top) the ENSO-based model described in \cref{sec:methods-enso} and (Bottom) the Markov chain modle described in \cref{sec:methods-markov}.
    Both models use a stationary parameterization.
    49 of the 50 sequences are plotted in gray, and one (randomly chosen) is plotted in blue for emphasis.
    The right hand panel shows the distribution over the historical and future periods for 500 sequences from the same model, visualized with a histogram and kernel density estimate.
    Horizontal line shows a threshold of \num{2500}, which is exceeded with probability \num{0.014133333333333333} by the NINO3 sequences and \num{0.013546666666666667} by the Markov sequences.\label{fig:stationary-sequences}
  }
\end{figure}

To incorporate secular variability into our model, we repeat the exercise but add a trend term to the data.
For the ENSO time series, this mean setting \(\gamma=0.02\) while for the Markov case this meant a trend term on the wet state only, so that \(\gamma_1=0.02\).
The resulting sequences are shown in \cref{fig:trend-sequences}.
The trend has an extremely large impact on the probability of an extremely high flood, but the underlying time series structure is generally preserved.
\begin{figure}[b]
  \includegraphics[width=\textwidth]{../../figs/trend_sequences.pdf}
  \caption{
    As \cref{fig:stationary-sequences} but a trend term has been added.
    Horizontal line shows a threshold of \num{2500}, which is exceeded with probability \num{0.023306666666666667} by the NINO3 sequences and \num{0.024493333333333332} by the Markov sequences.\label{fig:trend-sequences}
  }
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{../../figs/stationary_bias.pdf}
  \caption{The estimation bias for 1000* sequences generated by each of two stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{../../figs/stationary_variance.pdf}
  \caption{The estimation variance for 1000* sequences generated by each of two stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{../../figs/trend_bias.pdf}
  \caption{The estimation bias for 1000* sequences generated by each of two non-stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{../../figs/trend_variance.pdf}
  \caption{The estimation variance for 1000* sequences generated by each of two non-stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

From a decision making perspective, one could either design and build a project that gives us protection through the year 2100 considering both types of nonstationarities, or consider a sequence of projects every M years that incrementally add (or not) protection. How does the uncertainty and hence risk for these decisions manifest depending on the design level p, the duration M, and the uncertainty associated with the estimation of the \( f(Qp(t)) \) given a methodology used for prediction -- \eg{} the model chain using gcms and such and models fitted to a historical data length \(M\).
WHat is the implication for choices of p and M under different models for stationarity and uncertainty using some simple loss and cost  functions.

@@

\begin{figure}
  \includegraphics[width=\textwidth]{../../figs/enso.pdf}
  \caption{
    A \SI{2500}{year} sub-set of the synthetic time series described in \cref{sec:methods-enso}.\label{fig:enso-ts}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{../../figs/example_long.pdf}
  \caption{
    A single sequence of streamflow generated by the NINO3 function with (L) no trend and (R) \(\gamma=0.01\). 
    The sequence is fit to each f`it to three different fitting models (rows). 
    Generated with \(M=N=150\).\label{fig:example-long}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{../../figs/example_short.pdf}
  \caption{
    As \cref{fig:example-short} except \(N=30\).\label{fig:example-short}
  }
\end{figure}

>>>>

\input{header/macros.tex}
\input{header/packages.tex}

\title{Nonstationary, Quasi-Periodic Risk of Climate Extremes, and Decision Analysis}
\author[1,2]{James Doss-Gollin}
\author[1,2]{Upmanu Lall}
\author[1,2]{David Farnham}
\affil[1]{Columbia Water Center, Columbia University}
\affil[2]{Department of Earth and Environmental Engineering, Columbia University}

\input{header/biblatex.tex}
\addbibresource{library.bib}
\graphicspath{{../figs/}}

@@

\begin{document}
\maketitle
\RaggedRight{}
\begin{abstract}
  Hydroclimatic systems vary on many timescales, including regime-like variability and secular variability  \citep[``nonstationarity'';][]{Milly2008,Merz2014,Hodgkins2017}.
  This multi-scale variability greatly complicates the estimation of future flood sequences, leading to ``low confidence in projections of changes in fluvial floods'' \citep{IPCC2012}.
  Despite broad recognition of nonstationarity's importance, there has been little theoretical development of ideas for the design and operation of infrastructure that simultaneously consider the potential for predictability of secular change and regime behavior or low-frequency variability.
  Consequently, existing theories of flood risk assessment are well-suited to analysis of flood adaptation strategies with fixed project operation period \(M\), but not for analysis of sequential decision strategies.
  In this paper we illustrate some basic considerations of uncertainty and risk analysis associated with the sequential decisions that may need to be made for climate risk adaptation in a nonstationary adaptation.
  We simulate synthetic data and fit the resulting simulations in a Bayesian context using stationary and non-stationary models to show \(\ldots{}\).
  This is a first conceptual step to the development of a full sequential decision analysis approach under nonstationarity.
\end{abstract}

@@

The \nth{20} century saw the evolution of a nearly universal approach to the statistical analysis of climate extremes for use in risk analysis and mitigation studies \citep[\ie{} Bulletin 17-B for flood frequency analysis;][]{IACWD1982}.
In this framework event risk is assessed as a relationship between the event magnitude and its annual probability of exceedance.
Under the assumption that the historical or simulated record of events constitutes an independent and identically distributed process, and using an appropriate statistical inference model, an estimate of this relationship and often its uncertainty based on a finite \(N\) year record is developed.
This is then used as a measure for decision analysis for infrastructure, flood zoning and insurance instruments for event risk mitigation, typically using an expected net present value criteria applied to the costs and residual losses expected from the mitigation instrument.

Over time it has become clear that even small nonstationarity (temporal changes in dynamics of the underlying hydroclimatalogical process) can lead to significant trends in the estimated \(T\)-year return periods or probability of exceedance \( p_T \) \citep{Milly2008,Vogel2011,Salas2014,Merz2014,Obeysekera2016}.
This has led to a plethora of efforts aimed at identifying the nature of the nonstationarity and modeling it, \eg{} by introducing time trends on statistical parameters \citep{Strupczewski2001,Vogel2011,Obeysekera2014,Serinaldi2015}, climate covariates \citep{Delgado2014,Griffis2007,Silva2016,Sun2014,Hall2014}, latent variable or state-based modeling approaches \citep{Griffis2007,Sveinsson2005,Waylen1986}, or using only the last \(N\) years of the data.
Each model will necessarily give different estimates, particularly as the model is forecast further into the future.
Despite the wide suite of available models, the question of how one should use these nonstationarity estimates for the design of risk mitigation infrastructure or the pricing of insurance like instruments has been much less studied.

There are two kinds of nonstationairities of concern.
The first, secular change, may be related to anthropogenic climate change, river modification, or land use change \citep{Merz2014}.
The uncertainties associated with this are not purely informational; both nature and future human actions will determine the outcomes, and neither is obviously predictable.
The approach taken by many, including the IPCC, is to consider discrete scenarios for what humans may do with respect to greenhouse gas emissions and then use climate models with these scenarios to develop probabilistic statements as to the potential changes in the extremes.

Despite continuing improvements in physical understanding and computational capacity, there is cause for concern with dependence on using physics-based numerical models to project future climate risk.
Use of a ``model chain'' typically encompasses:
\begin{enumerate*}[label= (\roman*) ]
  \item emission scenario;
  \item general circulation model (GCM);
  \item downscaling;
  \item hydrological catchment model; and
  \item flood frequency analysis.
\end{enumerate*}
Given poor performance of flood projections relative to \nth{20} century data, ``bias correction methods'' whose applicability under extrapolation does not inspire confidence are typically applied \citep{Dankers2009,Ott2013,Merz2014,Dittes2017}.\footnote{Farnham et al 2018!}
A critical question in this regard is how to assess whether any structure designed using these projections may actually perform in the future relative to the proposed level of risk protection.
Would it be over- or under- designed?
Would the high degree of bias and uncertainty in its estimation lead to suboptimal investment strategies in risk mitigation?
These questions go far beyond what can be reliably answered at this time, and this type of epistemic uncertainty clearly grows as we look out into the distant future.

If a positive discount rate is used for the future, as is typical in public and private sector planning models,\footnote{CITE} the investment returns on major public and private investment may diminish substantially far into the future.
This consideration suggests that it may be more effective to understand the consequences for analysis focusing on shorter lived mitigation alternatives that may last for 1, 5, 10, 20, 30, or 50 years.
Thus, sequences of such smaller projects may in some cases better reflect climate adaptation goals, particularly given the possibilities that major climate change impacts in a paticular place may occur far into the future; that climate mitigation efforts may be successful; or that the needs of human society may change drastically over these horizons.

Consideration of these shorter time horizons brings up the concern associated with the second type of nonstationarity in climate: the quasi-periodic inter-annual, decadal and multidecadal modes of climate variability that have extensive spatial teleconnections and appear to influence the spatio-temporal expression of climate extremes in ways that violate the assumption of independent, identically distributed draws of events that is central to the classical analysis \citep[see][]{Merz2014,Serinaldi2015,Hoskins2015,Hannachi2017}.
Using a bootstrap approach with a long, stationary ENSO data series, \citet{Jain2001} demonstrated that the using the past \(K\) years to represent the next \(K\) years led to large surprise relative to when the data was fully randomized and the temporal structure removed.
Even for a very long sequence (\ie{} \(K=200\) years), the past was found to be a misleading representation of the future.
Such clustering in the occurrence of extreme rainfall and drought is manifest in the potential exposure of mining companies,\footnote{Bonnafous et al (2017a,b)} who face spatial as well as temporal clustering.

Clearly, in a nonstationary world the probability of exceedance is temporally variable over the project life.
In this paper we provide some initial insight into the bias and uncertainty associated with flood frequency estimation for projects as a function of the project planning period \(M\) and the length of the historical record \(N\).
This project planning period can be interpreted as the design life of a physical structure, or as the contract length of a financial instrument.
Our aim is not to identify the ``best'' model, but rather to build some theory as to how the underlying nature of the natural variability and nonstationarity and the choice of model used to model future flood risk map onto estimation bias and variance in the context of many different \(M\) and \(N\).
This estimation bias and variance may map directly onto the price of an insurance contract\footnote{cite} or the possibility of over/under-design of a flood control structure.
Though the specific examples target flood risk, this method and framework may be used for understanding the implications of uncertain projections of the future for climate hazards including drought, hurricanes, and wind.

We use a highly idealized modeling framework to generate sequences of annual-maximum floods, fit these sequences in a probabilistic framework to three flood models that incorporate time in different ways, and evalaluate the performance of these models for multiple observational record lengths \(N\) and project planning periods \(M\).

We generate synthetic flood sequences from two different simple conceptual models that represent chaotic dynamics and long memory.
For both cases, the expected value of the annual maximum flood flood follows a log-normal distribution, conditional on an expected value \(\mu(t)\) which depends on individual model:
\begin{equation} \label{eq:lognormal}
  \log Q(t) \big| \mu(t),\sigma(t) \sim \normal \qty(\mu(t), \sigma(t)).
\end{equation}.
For simplicitiy we assume a constant coefficient of variation:
\begin{equation}\label{eq:calc-sigma}
  \sigma(t) = 
  \begin{cases} 
    \alpha \mu(t) & \qqtext{if} \alpha \mu(t) \geq \sigma_\text{min} \\ 
    \sigma_\text{min} & \qqtext{else}
  \end{cases}
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.

The procedure to generate a sequence of streamflow is thus:
\begin{enumerate}
  \item Choose \(t_0\), \(M\), and \(N\) so that the time is \(\vb{t}=t_0-N+1, \ldots, t_0+M\).
  \item Use one of the models described below to simulate \(\mu(t)\)
  \item Use \cref{eq:calc-sigma} to calculate \(\sigma(t)\)
  \item For each time \(t \in \vb{t}\), draw \(Q(t)\) following \cref{eq:lognormal}.
\end{enumerate}
Though both approaches used to generate sequences \( \mu(t) \) use highly simplified models neglecting many important mechanisms, they enable us to generate many synthetic sequences of \(\qty[M + N]\) years which exhibit chaotic but organized quasi-periodic behavior (regime behavior or low-frequency variability) plus a trend term, thereby capturing some essential features of hydrological variability.
In \cref{sec:sequence-realistic} we show sequences of streamflow generated by both models for some choices of parameters and discuss the processes which they can reasonably represent.

\subsubsection{Synthetic NINO3 Model\label{sec:methods-nino}}

ENSO is the dominant mode of tropical climate variability, impacts flood risk around the world, and varies in a quasi-oscillatory manner with variability on the order of 3 to 7 years with further low-frequency modulation \citep{Sarachik2010}.
A \SI{100000}{year} integration of the Cane-Zebiak model \citep{Zebiak1987} was used to produce a monthly NINO3.4 index with stationary forcing as described in \citet{Ramesh2017}.
A subset of this data is shown in \cref{fig:enso-ts}.
To create an annual time series, we average the October-December values of the NINO3 index for each year.

To generate \(\mu(t)\), a sequence of length \(M+N\) is randomly chosen from the \SI{100000}{year} NINO3 sequence.
Next, \(\mu\) is calculated following by
\begin{equation}
  \mu(t) = \mu_0 + \beta x(t) + \gamma \qty(t - t_0)
\end{equation}
where parameter \(\gamma\) controls the magnitude of the secular trend (\(\gamma=0\) gives a stationary series).

This paper does not propose new models or suggest a general ``best'' model; instead it aims to describe the total uncertainty in estimates using easily methods which have been well studied in the literature.

In general, each historical \(N\)-year sequence of streamflow is fit to a statistical model, from which 1000\footnote{Check this parameter} sequences of future streamflow are generated using Monte Carlo simulation.
Three models are assessed for estimating future streamflow and are described in detail below.

\subsubsection{Stationary LN2 Model\label{sec:method-stationary}}

We first consider fitting a stationary model to the observed flood record, following classical assumptions of \iid sequences and stationarity.
We use the log-normal model (\cref{eq:lognormal}) for stationary flood distribution; while this model is less flexible than three-parameter models such as the GEV or LP3 \citep{Vogel1996}, it is easy to understand and interpret.
Since we know the true distribution of the flood sequences (we have generated them ourselves), using the log-normal for both generating and fitting enables interpretation of results as a lower bound on total uncertainty.

For each simulation, the first \(N\) years \(\qty(t=[t_0-N+1, \ldots, t_0])\) are treated as observations (denoted \(Q_{\text{hist}}\) for clarity).
Then, the parameters are fit in a Bayesian framework to fully represent the posterior uncertainty as implemented in the \texttt{stan} probabilistic programming language \citep{Carpenter2016}.
Weak (uninformative) priors are chosen; in the context of a synthetic data experiment using informative priors seems inappropriate but other studies\footnote{cite} have used scaling information and regional information to inform prior distributions and reduce posterior variance.
The full probability model is thus:
\begin{align}
  \log Q_\text{hist} & \sim \normal \qty(\mu, \sigma) \\
  \mu &\sim \normal \qty(0, 10) \\
  \sigma &\sim \normal \qty(0, 2)
\end{align}

\subsubsection{Trend LN2 Model\label{sec:method-trend}}

To incorporate secular change into a statistical model, many studies have used time parameterizations on certain parameters.
We consider a trend model where the data is log-normally distributed as \cref{eq:lognormal} with a linear time trend on both \( \mu \) and \( \sigma \):
\begin{align}
  \log Q_\text{hist} & \sim \normal \qty(\mu_0 + \beta_\mu \qty(t - t_0), \sigma_0 + \beta_\sigma \qty(t-t_0)) \\
  \mu_0 & \sim \normal \qty(0, 10) \\
  \beta_\mu & \sim \normal \qty(0, 0.1) \\
  \sigma_0 & \sim \normal \qty(0, 2) \\
  \beta_\sigma & \sim \normal \qty(0, 0.05)
\end{align}
As for the stationary case, we fit posterior samples in a Bayesian framework using \texttt{stan} and apply weak priors.
The influence of priors in this case is stronger than in the stationary case, and a reasonable choice of priors might more strongly favor \(\beta_mu=0\) and \(\beta_\sigma=0\), for example using the Horseshoe prior parameterization \citep{Piironen2016a}.
However, in the context of a synthetic data exercise it seems reasonable to apply priors that are at most weakly informative.

A Hidden Markov Model (HMM) is a statistical Markov model in which the system being modeled is assumed to be a Markov process with unobserved (\ie{} hidden) states \(S(t)\) \citep{Rabiner1986}.
The (unobserved) states evolve following a first-order Markov process (\(S(t)\) depends only on \(S(t-1)\)), and the observed variable (streamflow) depends only on the observed state.
As such, HMMs are a structured prediction method which extend general mixture models to sequences of data, where position in the sequence is relevant.
Because this structure allows the hidden state to represent hydroclimatological ``regimes'' \citep{Reinhold1982,Michelangeli1995,Merz2014}, HMMs have have been used extensively to model hydrological variables such as streamflow \citep[\ie{}][]{Bracken2016} or climate states such as ENSO.\footnote{Julian's work}

We fit observed streamflow sequences \(Q_\text{hist}\) using a HMM as implemented in the python \texttt{pomegranate} package \citep{Schreiber2016}.
The model is fit to the \(N\) observed years of data using the Baum-Welch algorithm, assuming that the data follow a log-normal distribution conditional only on the state.
This algorithm simultaneously estimates the transition matrix of the Markov process and the conditional parameters of each distribution.
To choose the number of hidden states, a log probability score is used.\footnote{Need to implement this?}
Future floods are estimated by simulating future states from the resulting matrix and then drawing floods log-normally using the estimated parameters.

\subsection{Code Availability}

\begin{figure}[b]
  \includegraphics[width=\textwidth]{stationary_sequences.pdf}  
  \caption{
    50 streamflow sequences from (Top) the ENSO-based model described in \cref{sec:methods-nino} and (Bottom) the Markov chain modle described in \cref{sec:methods-markov}.
    Both models use a stationary parameterization.
    49 of the 50 sequences are plotted in gray, and one (randomly chosen) is plotted in blue for emphasis.
    The right hand panel shows the distribution over the historical and future periods for 500 sequences from the same model, visualized with a histogram and kernel density estimate.
    Horizontal line shows a threshold of \num{2500}, which is exceeded with probability \num{0.014133333333333333} by the NINO3 sequences and \num{0.013546666666666667} by the Markov sequences.\label{fig:stationary-sequences}
  }
\end{figure}

To incorporate secular variability into our model, we repeat the exercise but add a trend term to the data.
For the ENSO time series, this mean setting \(\gamma=0.02\) while for the Markov case this meant a trend term on the wet state only, so that \(\gamma_1=0.02\).
The resulting sequences are shown in \cref{fig:trend-sequences}.
The trend has an extremely large impact on the probability of an extremely high flood, but the underlying time series structure is generally preserved.
\begin{figure}[b]
  \includegraphics[width=\textwidth]{trend_sequences.pdf}
  \caption{
    As \cref{fig:stationary-sequences} but a trend term has been added.
    Horizontal line shows a threshold of \num{2500}, which is exceeded with probability \num{0.023306666666666667} by the NINO3 sequences and \num{0.024493333333333332} by the Markov sequences.\label{fig:trend-sequences}
  }
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{stationary_bias.pdf}
  \caption{The estimation bias for 1000* sequences generated by each of two stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{stationary_variance.pdf}
  \caption{The estimation variance for 1000* sequences generated by each of two stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{trend_bias.pdf}
  \caption{The estimation bias for 1000* sequences generated by each of two non-stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{trend_variance.pdf}
  \caption{The estimation variance for 1000* sequences generated by each of two non-stationary generating functions (row) and fit by each of three fitting functions (columns) as a function of \(M\) and \(N\).\label{fig:stationary-bias}}
\end{figure}

From a decision making perspective, one could either design and build a project that gives us protection through the year 2100 considering both types of nonstationarities, or consider a sequence of projects every M years that incrementally add (or not) protection. How does the uncertainty and hence risk for these decisions manifest depending on the design level p, the duration M, and the uncertainty associated with the estimation of the \( f(Qp(t)) \) given a methodology used for prediction -- \eg{} the model chain using gcms and such and models fitted to a historical data length \(M\).
What is the implication for choices of p and M under different models for stationarity and uncertainty using some simple loss and cost  functions.

\section{Thanks}

The authors thank Nandini Ramesh of Columbia University for providing the synthetic NINO3 index from a \SI{1000000}{year} run of the Cane-Zebiak model as described in \citet{Ramesh2017}.
Kyle Johnson of Columbia University for his suggestions for improving reliability and documentation of the \texttt{floodsampling} package.

\section{Background Information and Conceptual Framing}

@@

@@

\begin{figure}
  \includegraphics[width=\textwidth]{enso.pdf}
  \caption{
    A \SI{2500}{year} sub-set of the synthetic time series described in \cref{sec:methods-nino}.\label{fig:enso-ts}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{example_long.pdf}
  \caption{
    A single sequence of streamflow generated by the NINO3 function with (L) no trend and (R) \(\gamma=0.01\). 
    The sequence is fit to each f`it to three different fitting models (rows). 
    Generated with \(M=N=150\).\label{fig:example-long}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{example_short.pdf}
  \caption{
    As \cref{fig:example-short} except \(N=30\).\label{fig:example-short}
  }
\end{figure}
